---
title: "FMG Goodness-of-Fit P-Values in lavaan"
subtitle: "Local demonstration and implementation notes"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
execute:
  warning: false
  message: false
---

## Purpose

This note demonstrates the local lavaan integration of the FMG
goodness-of-fit p-value methods from semTests. It is development
documentation, not package-facing Rd documentation.

The current implementation covers:

- one-model p-values through `lavaan(..., test = ...)`
- one-model p-values through `lavTest()`
- one-model p-values through `fitMeasures(..., fm.args = ...)`
- nested model comparisons through `lavTestLRT()`

The currently supported nested surface is focused on FMG eigenvalue-based
methods: `pall`, `all`, `peba`, `eba`, and `pols`.

## Changes vs. `origin/master`

The integration is deliberately narrow: one new source file plus small,
localized additions to existing files. No new runtime dependencies, no
NAMESPACE changes, no changes to the optimizer or to the `@test` slot shape.

| File | Change | Role |
|---|---|---|
| `R/lav_test_fmg.R` | **new**, ~950 lines | FMG parser, Imhof integrator, one-model + nested p-values, UGamma (biased/unbiased). |
| `R/lav_test.R` | +54 / -3 | Let FMG names pass `lav_test_rename()`; one dispatch branch in `lav_model_test()`. |
| `R/lav_test_LRT.R` | +123 | Early dispatch + `lav_test_lrt_fmg()` for nested comparisons. |
| `R/lav_options.R` | +4 | Whitelist FMG names in final `test` validation. |
| `R/lav_test_browne.R` | +6 / -2 | `chol()` -> `MASS::ginv()` fallback when the Browne NT fast-path projection matrix `A` is singular (grouped RLS). |
| `DESCRIPTION` | +2 | `Suggests: testthat` *(local only; to be removed before upstream PR)*. |

Total FMG-only footprint vs `origin/master`: 6 files, +1134 / -8 lines.

### What FMG does *not* touch

- No changes to NAMESPACE (all helpers are `@keywords internal`).
- No new runtime imports. The quadratic-form p-value uses a pure-R Imhof
  helper in `R/lav_test_fmg.R` rather than `CompQuadForm`.
- No changes to the estimator, the optimizer, or the fit-time control flow;
  FMG p-values are computed after `VCOV` is available, inside the existing
  `lav_model_test()` loop.
- No changes to the `lavobject@test` list shape. FMG entries use the standard
  `test`, `stat`, `stat.group`, `df`, `refdistr`, `pvalue` keys.

The Browne fast-path fallback is included because grouped RLS tests enter
`lav_test_browne_nt_fast()` for unrestricted models where `A` can be
rank-deficient. In the grouped Holzinger-Swineford RLS checks, `A` has rank 30
in a 60 by 60 matrix for each group, so plain `chol(A)` fails and the
Moore-Penrose solution `MASS::ginv(A) %*% b` is needed.

### Call-flow summary

`cfa(..., test = "peba4_rls")` walks the same path as any other test name:

1. `lav_options_set()` validates via `lav_test_rename()`; FMG names are
   recognized by `lav_test_fmg_is_fmg()` and passed through unchanged.
2. Estimation runs as usual. `VCOV` carries `E.inv`, `Delta`, `WLS.V`,
   `Gamma` as attributes.
3. `lav_lavaan_step14_test` calls `lav_model_test()`, which computes the
   standard chi-square first, then dispatches FMG names to `lav_test_fmg()`.
4. The result is stored in `@test` and is read by `summary()`,
   `lavInspect(fit, "test")`, `lavTest()`, and `fitMeasures()`.

`lavTestLRT(m0, m1, test = "pall_ml")` takes a parallel early-dispatch path
into `lav_test_lrt_fmg()`, which builds the standard `anova`-shaped data
frame and fills difference p-values from `lav_test_fmg_nested()`.

```{r setup}
#| include: false
find_repo_root <- function(path = getwd()) {
  path <- normalizePath(path, mustWork = TRUE)
  repeat {
    desc <- file.path(path, "DESCRIPTION")
    if (file.exists(desc)) {
      dcf <- read.dcf(desc)
      if (identical(unname(dcf[1, "Package"]), "lavaan")) {
        return(path)
      }
    }
    parent <- dirname(path)
    if (identical(parent, path)) {
      stop("Could not locate lavaan repository root.", call. = FALSE)
    }
    path <- parent
  }
}

repo_root <- find_repo_root()

invisible(capture.output(
  pkgload::load_all(repo_root, export_all = FALSE, quiet = TRUE)
))

fmt_tests <- function(x) {
  data.frame(
    test = names(x),
    chisq = vapply(x, `[[`, numeric(1), "stat"),
    df = vapply(x, `[[`, numeric(1), "df"),
    pvalue = vapply(x, `[[`, numeric(1), "pvalue"),
    row.names = NULL
  )
}

hs_model <- "
  visual  =~ x1 + x2 + x3
  textual =~ x4 + x5 + x6
  speed   =~ x7 + x8 + x9
"
```

## One-Model Tests

Fit a standard Holzinger-Swineford CFA model with `estimator = "MLM"` so
the robust Gamma-related ingredients are available.

```{r one-model-fit}
fit <- cfa(
  hs_model,
  data = HolzingerSwineford1939,
  estimator = "MLM"
)
```

### On demand with `lavTest()`

FMG tests can be requested after fitting. Suffixless names use the standard ML
chi-square statistic by default:

```{r lavtest-on-demand}
one_model_tests <- c(
  "peba4",
  "eba2",
  "pols2_ml",
  "pall",
  "all"
)

test_out <- lavTest(
  fit,
  test = one_model_tests,
  drop.list.single = FALSE
)

fmt_tests(test_out[one_model_tests])
```

Use lavaan's existing `scaled.test` option to switch the base chi-square
statistic to Browne's residual NT model-based statistic:

```{r lavtest-rls}
lavTest(
  fit,
  test = "peba4",
  scaled.test = "browne.residual.nt.model"
)
```

Use the existing `gamma.unbiased` lavaan option to request the unbiased Gamma
estimator:

```{r lavtest-unbiased}
fit_unbiased <- cfa(
  hs_model,
  data = HolzingerSwineford1939,
  estimator = "MLM",
  gamma.unbiased = TRUE
)

lavTest(
  fit_unbiased,
  test = "peba4",
  scaled.test = "browne.residual.nt.model"
)
```

### At fit time with `lavaan(..., test = ...)`

The same option pattern works at fit time. FMG results are stored in the `@test`
slot and are then available to summaries and downstream helpers.

```{r fit-time-tests}
fit_with_fmg <- cfa(
  hs_model,
  data = HolzingerSwineford1939,
  estimator = "MLM",
  test = "peba4",
  scaled.test = "browne.residual.nt.model",
  gamma.unbiased = TRUE
)

names(fit_with_fmg@test)

fmt_tests(fit_with_fmg@test["peba4"])
```

### Through `fitMeasures()`

`fitMeasures()` already has a selected-test mechanism. The FMG p-value is
available by passing the desired FMG test as `fm.args$standard.test`.

```{r fitmeasures}
fitMeasures(
  fit_with_fmg,
  fit.measures = c("chisq", "df", "pvalue"),
  fm.args = list(standard.test = "peba4")
)

fitMeasures(
  fit,
  fit.measures = c("chisq", "df", "pvalue"),
  fm.args = list(standard.test = "pols2_ml")
)
```

## Nested Model Tests

For nested comparisons, the currently supported FMG methods are `pall`, `all`,
`peba`, `eba`, and `pols`. Suffixless names again inherit `scaled.test` and
`gamma.unbiased` from the fitted models.

```{r nested-models}
hs_restricted <- "
  visual  =~ x1 + a*x2 + x3
  textual =~ x4 + a*x5 + a*x6
  speed   =~ x7 + a*x8 + x9
"

fit_restricted <- cfa(
  hs_restricted,
  data = HolzingerSwineford1939,
  estimator = "MLM",
  gamma.unbiased = TRUE
)

fit_unrestricted <- cfa(
  hs_model,
  data = HolzingerSwineford1939,
  estimator = "MLM",
  gamma.unbiased = TRUE
)
```

### Nested comparison

FMG nested tests use the Satorra (2000) UGamma projection. `method` must be
`"satorra.2000"` (or left at the default); `"satorra.bentler.2001"` is not
supported for FMG nested tests and errors.

```{r nested-2000}
lavTestLRT(
  fit_restricted,
  fit_unrestricted,
  method = "satorra.2000",
  test = "pall"
)
```

An RLS-based variant is selected through the fit-time options, not through
`method`:

```{r nested-rls}
fit_restricted_rls <- cfa(
  hs_restricted,
  data = HolzingerSwineford1939,
  estimator = "MLM",
  scaled.test = "browne.residual.nt.model",
  gamma.unbiased = TRUE
)

fit_unrestricted_rls <- cfa(
  hs_model,
  data = HolzingerSwineford1939,
  estimator = "MLM",
  scaled.test = "browne.residual.nt.model",
  gamma.unbiased = TRUE
)

lavTestLRT(
  fit_restricted_rls,
  fit_unrestricted_rls,
  method = "satorra.2000",
  test = "peba4"
)
```

### Grouped Nested Example

Grouped nested comparisons are also covered for the FMG eigenvalue methods.

```{r grouped-nested}
fit_group_restricted <- cfa(
  hs_model,
  data = HolzingerSwineford1939,
  group = "school",
  group.equal = "loadings",
  estimator = "MLM",
  gamma.unbiased = TRUE
)

fit_group_unrestricted <- cfa(
  hs_model,
  data = HolzingerSwineford1939,
  group = "school",
  estimator = "MLM",
  gamma.unbiased = TRUE
)

lavTestLRT(
  fit_group_restricted,
  fit_group_unrestricted,
  method = "satorra.2000",
  test = "pall"
)
```

Nested `pols` uses the same nested UGamma eigenvalues as the other FMG nested
methods and matches the semTests reference:

```{r nested-pols}
lavTestLRT(
  fit_restricted,
  fit_unrestricted,
  method = "satorra.2000",
  test = "pols2_ml"
)
```

## Test String Grammar

The user-facing API separates the three moving parts:

- FMG method: `test = "peba4"`, `test = "eba2"`, `test = "pall"`, etc.
- chi-square statistic: `scaled.test = "standard"` or
  `scaled.test = "browne.residual.nt.model"`.
- Gamma estimator: `gamma.unbiased = FALSE` or `TRUE`.

The semTests-style suffix grammar is still accepted as a compatibility layer:

```text
<method><parameter>_<ug?>_<chisq>
```

Examples:

- `peba4_rls`: pEBA with 4 blocks, biased Gamma, RLS statistic.
- `peba4_ug_rls`: pEBA with 4 blocks, unbiased Gamma, RLS statistic.
- `eba2_ml`: EBA with 2 blocks, biased Gamma, ML statistic.
- `pall_ug_ml`: penalized all-eigenvalue nested p-value, unbiased Gamma, ML statistic.

The chi-square statistic suffix is:

- `ml`: normal-theory ML chi-square.
- `rls`: Browne residual normal-theory/model-based RLS statistic.

The optional `ug` component requests the unbiased Gamma estimator.

## Implementation Map

The implementation is intentionally concentrated in a small number of files.

| File | Role |
|---|---|
| `R/lav_test_fmg.R` | FMG parser, one-model p-values, UGamma helpers, nested p-values. |
| `R/lav_test.R` | Allows FMG names in `lav_test_rename()` and dispatches from `lav_model_test()`. |
| `R/lav_options.R` | Allows FMG names through final `test` option validation. |
| `R/lav_test_LRT.R` | Adds a narrow FMG branch in `lavTestLRT()` for nested comparisons. |
| `DESCRIPTION` | Temporarily adds `testthat` for local parity tests. |
| `tests/testthat/` | Temporary parity tests against the local `semTests/` checkout. |

### One-model path

1. `lav_test_rename()` keeps FMG strings intact and validates them with
   `lav_test_fmg_is_fmg()`.
2. `lav_model_test()` detects FMG tests and calls `lav_test_fmg()`.
3. `lav_test_fmg()` resolves the requested ML or RLS statistic from
   `scaled.test`, resolves the Gamma estimator from `gamma.unbiased`,
   constructs UGamma, extracts eigenvalues, and computes the requested p-value.
4. `fitMeasures()` uses its existing selected-test mechanism through
   `fm.args$standard.test`; no separate named FMG fit measure is required.

### Nested path

1. `lavTestLRT()` detects FMG tests and dispatches to `lav_test_lrt_fmg()`.
2. `lav_test_lrt_fmg()` builds the standard lavaan `anova` table shape and
   fills the difference p-value from `lav_test_fmg_nested()`.
3. `lav_test_fmg_nested()` computes the nested chi-square difference and the
   eigenvalue reference distribution.
4. The Satorra-2000 projection with semTests-compatible group scaling is the
   only supported UGamma construction for FMG nested tests;
   `method = "satorra.bentler.2001"` is rejected because the scaled-and-shifted
   style FMG p-values could not be made to match a sensible semTests
   `method = "2001"` reference and the simpler `(U0 - U1) %*% Gamma`
   construction is intentionally left out.

## Validation Status

The temporary testthat suite compares the local implementation against the
local semTests checkout.

Covered one-model cases:

- ungrouped Holzinger-Swineford CFA
- grouped Holzinger-Swineford CFA
- meanstructure Holzinger-Swineford CFA
- `std`, `pall`, `all`, `peba`, `eba`, `pols`, `sb`, `ss`, and `sf` variants

Covered nested cases:

- ungrouped and grouped Holzinger-Swineford comparisons
- `pall`, `all`, `peba`, `eba`, and `pols` variants (Satorra-2000 UGamma only)

Package-level check status:

```text
R CMD build .
R CMD check lavaan_0.6-22.2469.tar.gz
Status: OK
```

## Open Decisions

- Whether to keep temporary testthat tests or convert them before an upstream
  pull request.
- Whether to document `fitMeasures(..., fm.args = list(standard.test = ...))`
  in Rd files or keep the first public documentation focused on `lavTest()`.
