---
title: "FMG Goodness-of-Fit P-Values in lavaan"
subtitle: "Local demonstration and implementation notes"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
execute:
  warning: false
  message: false
---

## Purpose

This note demonstrates the local lavaan integration of the FMG
goodness-of-fit p-value methods from semTests. It is development
documentation, not package-facing Rd documentation.

The current implementation covers:

- one-model p-values through `lavaan(..., test = ...)`
- one-model p-values through `lavTest()`
- one-model p-values through `fitMeasures(..., fm.args = ...)`
- nested model comparisons through `lavTestLRT()`

The currently supported nested surface is focused on FMG eigenvalue methods:
`pall`, `all`, `peba`, and `eba`. Nested `pols` is intentionally not supported.

```{r setup}
find_repo_root <- function(path = getwd()) {
  path <- normalizePath(path, mustWork = TRUE)
  repeat {
    desc <- file.path(path, "DESCRIPTION")
    if (file.exists(desc)) {
      dcf <- read.dcf(desc)
      if (identical(unname(dcf[1, "Package"]), "lavaan")) {
        return(path)
      }
    }
    parent <- dirname(path)
    if (identical(parent, path)) {
      stop("Could not locate lavaan repository root.", call. = FALSE)
    }
    path <- parent
  }
}

repo_root <- find_repo_root()

invisible(capture.output(
  pkgload::load_all(repo_root, export_all = FALSE, quiet = TRUE)
))

fmt_tests <- function(x) {
  data.frame(
    test = names(x),
    chisq = vapply(x, `[[`, numeric(1), "stat"),
    df = vapply(x, `[[`, numeric(1), "df"),
    pvalue = vapply(x, `[[`, numeric(1), "pvalue"),
    row.names = NULL
  )
}

hs_model <- "
  visual  =~ x1 + x2 + x3
  textual =~ x4 + x5 + x6
  speed   =~ x7 + x8 + x9
"
```

## One-Model Tests

Fit a standard Holzinger-Swineford CFA model with `estimator = "MLM"` so
the robust Gamma-related ingredients are available.

```{r one-model-fit}
fit <- cfa(
  hs_model,
  data = HolzingerSwineford1939,
  estimator = "MLM"
)
```

### On demand with `lavTest()`

FMG tests can be requested after fitting:

```{r lavtest-on-demand}
one_model_tests <- c(
  "peba4_rls",
  "peba4_ug_rls",
  "eba2_rls",
  "pols2_ml",
  "pall_rls",
  "all_rls"
)

test_out <- lavTest(
  fit,
  test = one_model_tests,
  drop.list.single = FALSE
)

fmt_tests(test_out[one_model_tests])
```

### At fit time with `lavaan(..., test = ...)`

The same tests can be computed while fitting. They are stored in the `@test`
slot and are then available to summaries and downstream helpers.

```{r fit-time-tests}
fit_with_fmg <- cfa(
  hs_model,
  data = HolzingerSwineford1939,
  estimator = "MLM",
  test = c("peba4_rls", "peba4_ug_rls", "pols2_ml")
)

names(fit_with_fmg@test)

fmt_tests(fit_with_fmg@test[c("peba4_rls", "peba4_ug_rls", "pols2_ml")])
```

### Through `fitMeasures()`

`fitMeasures()` already has a selected-test mechanism. The FMG p-value is
available by passing the desired FMG test as `fm.args$standard.test`.

```{r fitmeasures}
fitMeasures(
  fit,
  fit.measures = c("chisq", "df", "pvalue"),
  fm.args = list(standard.test = "peba4_rls")
)

fitMeasures(
  fit,
  fit.measures = c("chisq", "df", "pvalue"),
  fm.args = list(standard.test = "pols2_ml")
)
```

## Nested Model Tests

For nested comparisons, the currently supported FMG methods are `pall`, `all`,
`peba`, and `eba`. `pall_ug_ml` is the main recommended nested comparison.

```{r nested-models}
hs_restricted <- "
  visual  =~ x1 + a*x2 + x3
  textual =~ x4 + a*x5 + a*x6
  speed   =~ x7 + a*x8 + x9
"

fit_restricted <- cfa(
  hs_restricted,
  data = HolzingerSwineford1939,
  estimator = "MLM"
)

fit_unrestricted <- cfa(
  hs_model,
  data = HolzingerSwineford1939,
  estimator = "MLM"
)
```

### Method 2000

```{r nested-2000}
lavTestLRT(
  fit_restricted,
  fit_unrestricted,
  method = "satorra.2000",
  test = "pall_ug_ml"
)
```

### Method 2001

```{r nested-2001}
lavTestLRT(
  fit_restricted,
  fit_unrestricted,
  method = "satorra.bentler.2001",
  test = "peba4_ug_rls"
)
```

### Grouped Nested Example

Grouped nested comparisons are also covered for the FMG eigenvalue methods.

```{r grouped-nested}
fit_group_restricted <- cfa(
  hs_model,
  data = HolzingerSwineford1939,
  group = "school",
  group.equal = "loadings",
  estimator = "MLM"
)

fit_group_unrestricted <- cfa(
  hs_model,
  data = HolzingerSwineford1939,
  group = "school",
  estimator = "MLM"
)

lavTestLRT(
  fit_group_restricted,
  fit_group_unrestricted,
  method = "satorra.2000",
  test = "pall_ug_ml"
)
```

Nested `pols` is currently outside the supported surface:

```{r nested-pols-error, error=TRUE}
lavTestLRT(
  fit_restricted,
  fit_unrestricted,
  method = "satorra.2000",
  test = "pols2_ml"
)
```

## Test String Grammar

The test strings follow the semTests-style grammar:

```text
<method><parameter>_<ug?>_<chisq>
```

Examples:

- `peba4_rls`: pEBA with 4 blocks, biased Gamma, RLS statistic.
- `peba4_ug_rls`: pEBA with 4 blocks, unbiased Gamma, RLS statistic.
- `eba2_ml`: EBA with 2 blocks, biased Gamma, ML statistic.
- `pall_ug_ml`: penalized all-eigenvalue nested p-value, unbiased Gamma, ML statistic.

The chi-square statistic suffix is:

- `ml`: normal-theory ML chi-square.
- `rls`: Browne residual normal-theory/model-based RLS statistic.

The optional `ug` component requests the unbiased Gamma estimator.

## Implementation Map

The implementation is intentionally concentrated in a small number of files.

| File | Role |
|---|---|
| `R/lav_test_fmg.R` | FMG parser, one-model p-values, UGamma helpers, nested p-values. |
| `R/lav_test.R` | Allows FMG names in `lav_test_rename()` and dispatches from `lav_model_test()`. |
| `R/lav_options.R` | Allows FMG names through final `test` option validation. |
| `R/lav_test_LRT.R` | Adds a narrow FMG branch in `lavTestLRT()` for nested comparisons. |
| `DESCRIPTION` | Adds `CompQuadForm` for Imhof p-values; `testthat` is temporary for local parity tests. |
| `tests/testthat/` | Temporary parity tests against the local `semTests/` checkout. |

### One-model path

1. `lav_test_rename()` keeps FMG strings intact and validates them with
   `lav_test_fmg_is_fmg()`.
2. `lav_model_test()` detects FMG tests and calls `lav_test_fmg()`.
3. `lav_test_fmg()` selects the requested ML or RLS statistic, constructs
   UGamma, extracts eigenvalues, and computes the requested p-value.
4. `fitMeasures()` uses its existing selected-test mechanism through
   `fm.args$standard.test`; no separate named FMG fit measure is required.

### Nested path

1. `lavTestLRT()` detects FMG tests and dispatches to `lav_test_lrt_fmg()`.
2. `lav_test_lrt_fmg()` builds the standard lavaan `anova` table shape and
   fills the difference p-value from `lav_test_fmg_nested()`.
3. `lav_test_fmg_nested()` computes the nested chi-square difference and the
   eigenvalue reference distribution.
4. Method `"2000"` uses the Satorra-2000 projection and semTests-compatible
   group scaling for the full UGamma matrix.
5. Method `"2001"` uses `(U0 - U1) %*% Gamma`.

## Validation Status

The temporary testthat suite compares the local implementation against the
local semTests checkout.

Covered one-model cases:

- ungrouped Holzinger-Swineford CFA
- grouped Holzinger-Swineford CFA
- meanstructure Holzinger-Swineford CFA
- `std`, `pall`, `all`, `peba`, `eba`, `pols`, `sb`, `ss`, and `sf` variants

Covered nested cases:

- ungrouped and grouped Holzinger-Swineford comparisons
- methods `"2000"` and `"2001"`
- `pall`, `all`, `peba`, and `eba` variants

Package-level check status:

```text
R CMD build .
R CMD check lavaan_0.6-22.2469.tar.gz
Status: OK
```

## Open Decisions

- Whether to keep temporary testthat tests or convert them before an upstream
  pull request.
- Whether to document `fitMeasures(..., fm.args = list(standard.test = ...))`
  in Rd files or keep the first public documentation focused on `lavTest()`.
- Whether nested `pols` should ever be added. It is currently excluded because
  nested support is scoped to PALL, ALL, pEBA, and EBA.
